### 2021-07-09 rad-icfp-2021/correctness.tex
some small fixes

\begin{definition}[Mutlicat from concat]
    We define $\CSource$ to be the multicategory with the same objects as Source and 
    $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$.
    The composition of $f_i:\underline{A}_i\to B_i$ with $g:\underline{B}\to C$ is given 
    $f_1\concatcomp\ldots f_n;(u_{\underline{B}_1}\concatcomp A_1 \concatcomp \ldots \concatcomp u_{\underline{B}_n}\concatcomp A_n);g$.
    In other words each term $f_i$ forgets about its output $\underline{B}_i$, and then we use the composition in the concategory.
\end{definition}

@@

\begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.\\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
    !$\Dsynrevsymbol_{\rho}$(op1)=<op1,$\pi_{last}\circ$ J$^T$op1>!,\\ 
    !$\Dsynrevsymbol_{\rho}$(op2)=<op2,$\pi_{last}\circ$ J$^T$op2>!,\\
    !$\Dsynrevsymbol_{\rho}$(map2 (x,y.e))=<map2 (x,y.e),$\pi_{last}\circ$ J$^T$map2 (x,y.e)>!,\\
    !$\Dsynrevsymbol_{\rho}$(reduce$_{x,y.e_1,e_2}$)=<reduce$_{x,y.e_1,e_2}$,$\pi_{last}\circ$ J$^T$reduce$_{x,y.e_1,e_2}$>!, 
    where $\pi_{last}$ is the projection to the last element (the continuation variable).
    It naturally extends on sequential composition. It does not automatically extend to a multi-functor between concategories because \\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$] $\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]$\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.
    However there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
    Internally as a lambda term, it is given by \\
    $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
    Here, $\widehat{+}$ is the reverse-derivative of the copy map which is known to be fanout. 
    It is not surprising to see it appear as $\Dsynrevsymbol_{\rho}$ is a semantic functor, and it does not need to be efficient. 
\end{definition}

>>>>

\begin{definition}[Mutlicategory from concat]
    We define $\CSource$ to be the multicategory with the same objects as Source and\\ 
    $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$.
    The composition of $f_i:\underline{A}_i\to B_i$ with $g:\underline{B}\to C$ is given 
    $f_1\concatcomp\ldots\concatcomp f_n;(u_{\underline{B}_1}\concatcomp A_1 \concatcomp \ldots \concatcomp u_{\underline{B}_n}\concatcomp A_n);g$.
    In other words each term $f_i$ forgets about its output $\underline{B}_i$, and then we use the composition in the concategory.
\end{definition}

@@

\begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.\\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
    !$\Dsynrevsymbol_{\rho}$(op1)=<op1,$\pi_{last}\circ$ J$^T$op1>!,\\ 
    !$\Dsynrevsymbol_{\rho}$(op2)=<op2,$\pi_{last}\circ$ J$^T$op2>!,\\
    !$\Dsynrevsymbol_{\rho}$(map2 (x,y.e))=<map2 (x,y.e),$\pi_{last}\circ$ J$^T$map2 (x,y.e)>!,\\
    !$\Dsynrevsymbol_{\rho}$(reduce$_{x,y.e_1,e_2}$)=<reduce$_{x,y.e_1,e_2}$,$\pi_{last}\circ$ J$^T$reduce$_{x,y.e_1,e_2}$>!, 
    where $\pi_{last}$ is the projection to the last element (the continuation variable).
    It naturally extends on sequential composition. 
    It does not automatically extend to a multi-functor between concategories because \\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]!{$\concatcomp$}![A$_{k+1}$,$\ldots$,A$_n$] $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
    However there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
    Internally as a lambda term, it is given by \\
    $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
    Here, $\widehat{+}$ is the reverse-derivative of the copy map which is known to be fanout. 
    It is not surprising to see it appear as $\Dsynrevsymbol_{\rho}$ is a semantic functor, and it does not need to be efficient. 
\end{definition}

### 2021-07-09 rad-icfp-2021/relwork.tex
some small fixes

An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
This work is purely first-order, does not have arrays and uses a non-standard semantics. 
Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.

@@

Relatively recent work in AD (2008-2020):
\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}

>>>>

An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
This work is purely first-order, does not have arrays and uses a non-standard semantics. 
Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
Instead, they follow the principle of unrolling the recursive calls and evaluating the conditional before differentiating, 
which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
This is a known limitation in TensorFlow \cite{abadi2016tensorflow}. %which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.

@@

% Relatively recent work in AD (2008-2020):
% \cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}
